---
title: "Advanced R: Object Oriented Programming"
subtitle: "Applying R to Lifestyle and Brain Health Research"
author: "Brian C. Helsel, PhD"
institute: "University of Kansas Medical Center"
date: today
format:
  revealjs:
    theme: [default, "../../styles/styleReveal.scss"]
    slide-number: true
    preview-links: auto
    logo: https://brand.ku.edu/sites/brand/files/styles/wide_col_xl_6/public/images/2020/KUMarksLogo.png
    footer: "R for Lifestyle and Brain Health (R-LAB)"
    transition: slide
    background-transition: fade
    highlight-style: github
    code-line-numbers: true
    code-block-bg: true
    incremental: false
    date-format: long
    embed-resources: true
execute: 
  echo: true
  eval: false
---

## Introduction

The first big idea behind **metaprogramming** is that *code is data*. You
can capture and compute on code like any other type of data. For example,
the `rlang::expr` function can return exactly what you pass in.


```{r}

rlang::expr(mean(x, na.rm = TRUE))
#> mean(x, na.rm = TRUE)

rlang::expr(10 + 100 + 1000)
#> 10 + 100 + 1000

```

Captured code is called an **expression**, a collective term for any of 
four types (call, symbol, constant, or pairlist). The **rlang::expr** 
allows you to capture code you type, but you need `rlang::enexpr` to 
capture code passed to a function.

```{r}

capture_it <- function(x, method) {
  if (method == "expr") {
    rlang::expr(x)
  } else if (method == "enexpr") {
    rlang::enexpr(x)
  }
}

capture_it(x = a + b + c, method = "expr")
#> x

capture_it(x = a + b + c, method = "enexpr")
#> a + b + c

```

Once you capture an expression, you can inspect and modify it like lists
using `[[` and `$`.

```{r}

f <- rlang::expr(f(x = 1, y = 2))

f$z <- 3
f
#> f(x = 1, y = 2, z = 3)

# The first argument is in the second position as the first element
# is the function to be called (i.e., f).

f[[2]] <- NULL
f
#> f(y = 2, z = 3)

```

## Code is a Tree

More complex manipulations with expressions require an understanding of
their structure. Nearly every programming language represents code as
an abstract syntax tree (AST). In R, you can inspect and manipulate 
the AST.

```{r}
#| eval: true

lobstr::ast(1 + 2 * 3)

lobstr::ast(paste("Mean:", mean(1:100), "SD: ", sd(1:100)))

```

## Code can Generate Code {.pattern-dots}

You can use `rlang::call2` and unquoting to use code to create new trees.
The **rlang::call2** function constructs a function call from its 
components: the function to the call and the arguments to call it with.

```{r}

rlang::call2("+", 1, rlang::call2("*", 2, 3))
#> 1 + 2 * 3

```

You can build complex code trees by combining simpler ones with a template.
The **expr** and **enexpr** functions have built-in support for this idea
via `!!`, the unquote operator (pronounced bang-bang).

```{r}

xx <- rlang::expr(x + y)
yy <- rlang::expr(y + y)

# !!xx inserts the code tree stored in xx into the expression

rlang::expr(!!xx / !!yy)
#> (x + y)/(y + y)

```

Unquoting is even more useful inside a function. For example, here is an
expression to calculate the coefficient of variation.

```{r}

cv <- function(var) {
  var <- rlang::enexpr(var)
  rlang::expr(sd(!!var) / mean(!!var))
}

cv(x)
#> sd(x)/mean(y)

```

## Evaluation Runs Code

You can evaluate (i.e., execute or run) an expression with `base::eval`.
The **eval** function takes two arguments: the expression and environment.
Omitting the environment causes **eval** to use the current one.

```{r}

eval(rlang::expr(x + y), rlang::env(x = 1, y = 10))
#> [1] 11

```


## Customizing Evaluation with Functions

You can also bind names to functions when evaluating which can allow you
to override the behavior of existing functions. For example, we can
override the `*` and `+` function to work with strings.

```{r}

add_string <- function(x) {
  e <- rlang::env(
    rlang::caller_env(),
    `+` = function(x, y) paste0(x, y)
  )
  eval(rlang::enexpr(x), e)
}

name <- "Brian"
add_string("Hello " + name)
#> [1] "Hello Brian"

```

The **dplyr** package uses this idea to run code in an environment to
generate SQL for execution in a remote database.

```{r}

con <- DBI::dbConnect(RSQLite::SQLite(), filename = ":memory:")
mtcars_db <- dplyr::copy_to(con, mtcars)

mtcars_db |>
  dplyr::filter(cyl > 2) |>
  dplyr::select(mpg:hp) |>
  head(10) |>
  dplyr::show_query()

#> <SQL>
#> SELECT `mpg`, `cyl`, `disp`, `hp`
#> FROM `mtcars`
#> WHERE (`cyl` > 2.0)
#> LIMIT 10

DBI::dbDisconnect(con)

```

## Customizing Evaluation with Data

You can modify evaluation to look for variables in a data frame instead
of an environment. This idea powers the base **subset** and **transform**
functions, as well as many tidyverse functions like **ggplot2::aes** and
**dplyr::mutate**. We can use `rlang::eval_tidy` which takes an 
expression, environment, and a data mask.

```{r}

df <- data.frame(x = 1:5, y = 6:10)
rlang::eval_tidy(rlang::expr(x + y), df)
#> [1] 7 9 11 13 15

```

We can expand upon this idea by wrapping this pattern into a function
using **enexpr** to get a similar function to `base::with`.

```{r}

with2 <- function(df, expr) {
  rlang::eval_tidy(rlang::enexpr(expr), df)
}

with2(df, x + y)
#> [1] 7 9 11 13 15

```