---
title: "Advanced R: Function Operators"
subtitle: "Applying R to Lifestyle and Brain Health Research"
author: "Brian C. Helsel, PhD"
institute: "University of Kansas Medical Center"
date: today
format:
  revealjs:
    theme: [default, "../../../styles/styleReveal.scss"]
    slide-number: true
    preview-links: auto
    logo: https://brand.ku.edu/sites/brand/files/styles/wide_col_xl_6/public/images/2020/KUMarksLogo.png
    footer: "R for Lifestyle and Brain Health (R-LAB)"
    transition: slide
    background-transition: fade
    highlight-style: github
    code-line-numbers: true
    code-block-bg: true
    incremental: false
    date-format: long
    embed-resources: true
execute: 
  echo: true
  eval: false
  freeze: auto
---

# Introduction

A **function operator** is a function that takes one (or more)
functions as input and returns a function as output. Function 
operators are function factories that take a function as an input 
and are typically paired with functionals.

## A Basic Example {.pattern-dots}

```{r}
chatty <- function(f) {
  force(f)
  function(x, ...) {
    res <- f(x, ...)
    cat("Processing ", x, "\n", sep = "")
    Sys.sleep(0.5)
    res
  }
}

purrr::map_dbl(c(3, 2, 1), chatty(function(x) x^2))
#> Processing 3
#> Processing 2
#> Processing 1
#> [1] 9 4 1
```

## Existing Function Operators

Two existing function operators that can help solve common recurring problems:

- Capturing errors with **purrr::safely**
- Caching computations with **memoise::memoise**

## Capturing Errors

One advantage of using **for** loops is being able to capture results up to
a failure. A functional does not return any output which makes it difficult to
figure out the problem.

```{r}
x <- list(
  c(0.51, 0.17, 0.72),
  c(0.06, 0.78, 0.43),
  c(0.89, 0.79, 0.50),
  "This is a problem!"
)

out <- rep(NA_real_, length(x))
for (i in seq_along(x)) {
  out[[i]] <- sum(x[[i]])
}
#> Error in `sum()`: invalid 'type' (character) of argument

out
#> [1] 1.40 1.27 2.18   NA

purrr::map_dbl(x, sum)
#> Error in `purrr::map_dbl()`: In index: 4.
#> Caused by error: ! invalid 'type' (character) of argument
```

The function operator `safely()` from the **purrr** package that transforms a
function to turn errors into data. A function transformed by **safely** always
returns a list with two elements: result and error.

```{r}
safe_sum <- purrr::safely(sum)
safe_sum
#> function (...)
#> capture_error(.f(...), otherwise, quiet)
#> <bytecode: 0x13d0c3be0>
#> <environment: 0x13dff12c8>

str(safe_sum(x[[1]]))
#> List of 2
#>  $ result: num 1.4
#>  $ error : NULL

str(safe_sum(x[[4]]))
#> List of 2
#>  $ result: NULL
#>  $ error :List of 2
#>   ..$ message: chr "invalid 'type' (character) of argument"
#>   ..$ call   : language .Primitive("sum")(..., na.rm = na.rm)
#>   ..- attr(*, "class")= chr [1:3] "simpleError" "error" "condition"
```

Using **safely** with a functional returns all of the results in a list,
but we can **transpose** it to get a list of results and errors.

```{r}
out <- purrr::map(x, purrr::safely(sum))
str(out)
#> List of 4
#>  $ :List of 2
#>   ..$ result: num 1.4
#>   ..$ error : NULL
#>  $ :List of 2
#>   ..$ result: num 1.27
#>   ..$ error : NULL
#>  $ :List of 2
#>   ..$ result: num 2.18
#>   ..$ error : NULL
#>  $ :List of 2
#>   ..$ result: NULL
#>   ..$ error :List of 2
#>   .. ..$ message: chr "invalid 'type' (character) of argument"
#>   .. ..$ call   : language .Primitive("sum")(..., na.rm = na.rm)
#>   .. ..- attr(*, "class")= chr [1:3] "simpleError" "error" "condition"

out <- purrr::transpose(purrr::map(x, purrr::safely(sum)))
str(out)

#> List of 2
#>  $ result:List of 4
#>   ..$ : num 1.4
#>   ..$ : num 1.27
#>   ..$ : num 2.18
#>   ..$ : NULL
#>  $ error :List of 4
#>   ..$ : NULL
#>   ..$ : NULL
#>   ..$ : NULL
#>   ..$ :List of 2
#>   .. ..$ message: chr "invalid 'type' (character) of argument"
#>   .. ..$ call   : language .Primitive("sum")(..., na.rm = na.rm)
#>   .. ..- attr(*, "class")= chr [1:3] "simpleError" "error" "condition"
```

We can now easily find the results that worked or the inputs that failed. This
approach could also work for fitting a linear model to multiple data sets and
looking back on the results that failed.

```{r}
ok <- purrr::map_lgl(out$error, is.null)

x[!ok]
#> [[1]]
#> [1] "This is a problem!"

unlist(out$result[ok])
#> [1] 1.40 1.27 2.18
```

## Other Function Operators in purrr

- Return a default value when there is an error with **possibly**
- Turn output, messages, and warning side-effects into output, message, and warning components of the output with **quietly**
- Automatically execute *browser* inside the function when there is an error with **auto_browser**

## Caching Computations

The function **memoise** from the memoise package will remember previous 
inputs and return cached results. Memoisation is an example of the classic
tradeoff between memory and speed. A memoised funcition can run faster, but
it takes up more memory as it stores all the prior inputs and outputs.


```{r}
slow_function <- function(x) {
  Sys.sleep(1)
  set.seed(123)
  x * 10 * runif(1)
}

system.time(print(slow_function(1)))
#> [1] 2.875775
#>    user  system elapsed
#>   0.003   0.002   1.003

fast_function <- memoise::memoise(slow_function)

# Elapsed time is the same as slow function the first time
system.time(print(fast_function(1)))
#> [1] 2.875775
#>    user  system elapsed
#>   0.002   0.003   1.002

# Elapsed time is much faster the second time
system.time(print(fast_function(1)))
#> [1] 2.875775
#>    user  system elapsed
#>   0.021   0.001   0.007
```

A realistic example of using memoisation is computing the Fibonacci series. The
Fibonacci series is defined recursively with the first two values defined as
$f(0) = 0$, $f(n) = 1$, adn then $f(n) = f(n - 1) + f(n - 2)$ for any positive
integer. A naive version is slow because it needs to compute multiple Fibonacci
values within the function.

```{r}
fib <- memoise::memoise(function(n) {
  if (n < 2) {
    return(1)
  }
  fib(n - 2) + fib(n - 1)
})

system.time(fib(23))
#>  user  system elapsed
#> 0.011   0.002   0.005

system.time(fib(23))
#> user  system elapsed
#>  0       0       0
```

::: {.highlight-box}

This illustrates dynamic programming, where a complex problem is divided into 
smaller subproblems, and storing their solutions significantly enhances efficiency.

:::

## Creating Your Own Function Operator

Imagine you have a named vector of URLs and you want to download each one to your 
computer. It is simple with **purrr::walk2** and **utils::file.download()**.

```{r}
urls <- c(
  "adv-r" = "https://adv-r.hadley.nz",
  "r4ds" = "https://r4ds.had.co.nz"
)

paths <- paste0(tempdir(), names(urls), ".html")

purrr::walk2(urls, paths, utils::download.file, quiet = TRUE)
```

This works great for a few URLs, but as the vector gets longer, you might want to
add a small delay to avoid overloading the server and display a `.` every few
URLs to know the function is still working. You can do this with a **for** loop,
but the code is harder to read and the components are not reusable in new 
situations.

```{r}
for (i in seq_along(urls)) {
  Sys.sleep(0.1)
  if (i %% 10 == 0) {
    cat(".")
  }
  utils::download.file(urls[[i]], paths[[i]])
}
```

A function operator overcomes the limitations of a *for* loop and we can create
`delay_by` and `dot_every` functions that can be reused.

```{r}
delay_by <- function(f, amount) {
  force(f)
  force(amount)
  function(...) {
    Sys.sleep(amount)
    f(...)
  }
}

dot_every <- function(f, n) {
  force(f)
  force(n)
  i <- 0
  function(...) {
    i <<- i + 1
    if (i %% n == 0) {
      cat(".")
    }
    f(...)
  }
}

purrr::walk2(
  urls,
  paths,
  utils::download.file |> dot_every(10) |> delay_by(0.1),
  quiet = TRUE
)
```