---
title: "Advanced R: Conditions"
subtitle: "Applying R to Lifestyle and Brain Health Research"
author: "Brian C. Helsel, PhD"
institute: "University of Kansas Medical Center"
date: today
format:
  revealjs:
    theme: [default, "../../styles/styleReveal.scss"]
    slide-number: true
    preview-links: auto
    logo: https://brand.ku.edu/sites/brand/files/styles/wide_col_xl_6/public/images/2020/KUMarksLogo.png
    footer: "R for Lifestyle and Brain Health (R-LAB)"
    transition: slide
    background-transition: fade
    highlight-style: github
    code-line-numbers: true
    code-block-bg: true
    incremental: false
    date-format: long
    embed-resources: true
execute: 
  echo: true
  eval: false
---

## Introduction

Provides a paired set of tools that allows the author of a function to indicate
that something unusual is happening and a user to deal with it. The author
**signals** conditions with functions like `stop()` (for errors), `warning()`
(for warnings), and `message()` (for messages). 

The user can handle the signals with functions like `tryCatch()` and `withCallingHandlers()`.

## Signalling conditions

- **Errors** indicate that there is no way for a function to continue and execution must stop
- **Warnings** show that something has gone wrong but the function could at least partially recover
- **Messages** inform users that some action has been performed on their behalf

Conditions are displayed prominently with errors always starting with **Error**
and warnings starting with **Warning**.

```{r}

stop("This is an error")
#> Error: This is an error

```


```{r}
#| eval: true
#| warning: true
#| message: true

warning("Here is a warning...")

message("and a message")

```

## Errors

Errors are signalled, or **thrown** using `stop()` Adding `call. = FALSE` removes the function call which is 
typically not useful and can be obtained from **traceback**. This can be accomplished automatically with
`rlang::abort`.

```{r}

f <- function() g()
g <- function() h()
h <- function() stop("Here is an error!", call. = FALSE)

f()
#> Error: Here is an error!

```

::: {.tip-box}

The best error messages tell you what is wrong and provides direction toward fixing the problem. Writing good
error messages is challenging because it is hard to imagine how the user is thinking incorrectly about your code.

:::

You can visit the <a href = "http://style.tidyverse.org/errors.html">tidyverse style guide</a> for helpful tips in writing error messages.

## Warnings

Warnings signal that something has gone wrong, but the code recovered and was able to continue. You can
have multiple warnings within a function. 

```{r}

fw <- function() {
  cat("1\n")
  warning("Warning #1", call. = FALSE)
  cat("2\n")
  warning("Warning #2", call. = FALSE)
}

fw()

#> 1
#> 2
#> Warning messages:
#> 1: Warning #1
#> 2: Warning #2

```

Warnings are cached and printed only when a function exists, but you can control this behavior with the `warn` option.

- `options(warn = 0)` is the default behavior and prints warnings at the end.
- `options(warn = 1)` will make the warnings appear immediately.
- `options(warn = 2)` will trun the warnings into errors (helps with debugging a warning).

It is difficult to know when to use warnings. Warnings are often missed if there is a lot of other output and restraining use
of warnings is advised. Warnings are appropriate when:

- You deprecate a function and want to allow older code to run but encourage users to switch to the newer function
- You are reasonable certain you can recover but not fix a problem

## Messages

Provide information to let the user know you have done something on their behalf. Good messages provide useful information
so the user knows what is happening, but not too many details that the user feels overwhelmed. Messages are recommended when:

- Providing a default argument in a function that requires some computation and you want to let the user know what value was used
(e.g., ggplot2 reports the number of bins if binwidth is missing)
- Using functions that are called for their side-effects and would otherwise be silent (e.g., writing files to disk or calling a web API)
- Starting a long, multi-step running process
- Providing a message when your package is loaded (e.g., `.onAttach()`)

::: {.tip-box}

The results of the `message()` and `cat()` functions are similar but their purposes are different. Use **cat** when the 
*user asks* for a message to be printed to the console and **message** when the *developer elects* for a message to
be printed.

:::

## Ignoring Conditions

A simple way to handle conditions in R is to ignore them. This can be done using **try** for errors, **suppressWarnings**
for warnings, and **suppressMessages** for messages.

The `try()` function allows execution to continue after an error has occurred.

```{r}

f1 <- function(x) {
  try(log(x))
  10
}

f1()

#> Error in log(x) : argument "x" is missing, with no default
#> [1] 10

```

By doing assignment within the call, you can define a default value for use if the code does not work. This works
because the argument is evaluated in the calling environment and not the function.

```{r}

default <- NULL

try(default <- read.csv("this-file-does-not-exist.csv"), silent = TRUE)

```

Both `suppressWarnings()` and `suppressMessages()` suppress all warnings and messages.

```{r}

suppressWarnings({
  warning("This might be a problem!")
  1
})
#> [1] 1

suppressMessages({
  message("Greetings!")
  2
})
#> [1] 2

```

## Handling Conditions

Each signalling condition has a default behavior:

- Errors stop execution and return to the top level
- Warnings are captured and displayed in aggregate
- Messages are immediately shown

Condition **handlers** allow us to temporarily override or supplement the default behavior.
We can use `tryCatch` and `withCallingHandlers` to register handlers.

```{r}

tryCatch(
  error = function(condition) {
    # code to run when an error is thrown
  },
  # code to run while handlers are active
)

withCallingHandlers(
  warning = function(condition) {
    # code to run when a warning is signalled
  },

  message = function(condition) {
    # code to run when a message is signalled
  },

  # code to run while handlers are active
)

```

## Condition Objects

Viewing a condition object that is signalled can be done with `rlang::catch_cnd`. Built-in
conditions have two elements:

- A **message** containing the text that should be displayed to the user
- A **call** that triggered the condition

```{r}

cnd <- rlang::catch_cnd(stop("Here is an error!"))
str(cnd)

#> List of 2
#>  $ message: chr "Here is an error!"
#>  $ call   : language force(expr)
#>  - attr(*, "class")= chr [1:3] "simpleError" "error" "condition"

# Extract the message
conditionMessage(cnd)

# Extract the call
conditionCall(cnd)

```

## Exiting Handlers

The function `tryCatch` can be used to handle error conditions. You can override the default
error behavior (e.g., return NA instead of an error) and code executes normally if no error 
exists.

```{r}

f2 <- function(x) {
  tryCatch(
    error = function(cnd) NA,
    log(x)
  )
}

f2("x")
#> [1] NA

f2(10)
#> [1] 2.302585

```