---
title: "Introduction to R Programming"
subtitle: "A Statistical Computing Environment"
author: "Your Name"
institute: "Your Institution"
date: today
format:
  revealjs:
    theme: [default, "../../styles/styleReveal.scss", customReveal.scss]
    slide-number: true
    preview-links: auto
    logo: https://www.r-project.org/logo/Rlogo.svg
    footer: "R Programming Workshop"
    transition: slide
    background-transition: fade
    highlight-style: github
    code-line-numbers: true
    code-block-bg: true
    incremental: false
    date-format: long
    embed-resources: true
---

## What is R? {.title-slide}

::: {.highlight-box}
**R is a language and environment for statistical computing and graphics**

:::

. . .

**Key Strengths:**

- Extensive package ecosystem (CRAN)
- Powerful data visualization tools
- Active research community
- Open source and free

## Getting Started {.pattern-dots}

::: {.columns}
::: {.column width="60%"}
**Installation Requirements:**

- **Base R**: Core language
- **RStudio**: Integrated development environment
- **Packages**: Extended functionality

::: {.tip-box}
💡 **Tip**: Always update R and packages regularly for latest features
:::
:::

::: {.column width="40%"}
**Essential Packages:**

| Package | Purpose |
|---------|---------|
| tidyverse | Data manipulation |
| ggplot2 | Visualization |
| dplyr | Data wrangling |
| readr | Data import |
:::
:::

## R Data Types and Structures {.pattern-grid}

::: {.code-box}
**Fundamental Data Types**
:::

```{r}
#| echo: true

# Numeric data
numeric_var <- 42.7
class(numeric_var)

# Character strings  
text_var <- "Hello, R!"
nchar(text_var)

# Logical values
logical_var <- TRUE
is.logical(logical_var)

# Factors for categories
factor_var <- factor(c("Low", "Medium", "High"))
levels(factor_var)

```

## Data Structures in R

**Vector Operations:**

```{r}
#| echo: true
#| eval: true

# Creating vectors
numbers <- c(1, 5, 10, 15, 20)
names <- c("Alice", "Bob", "Charlie")

# Vector arithmetic (vectorized operations)
numbers * 2
```

. . .

::: {.highlight-box}
**Data Frames - The Heart of R**

Data frames are R's primary data structure for rectangular data, similar to spreadsheets or SQL tables.
:::

```{r}
#| echo: true
#| eval: true

# Creating a data frame
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 35),
  Score = c(85, 92, 78)
)
head(df, 2)
```

## Data Import and Export

> **Best Practice**: Always examine your data structure with `str()`, `summary()`, and `head()` after importing

::: {.columns}
::: {.column width="50%"}
**Reading Data:**

```{r}
#| echo: true
#| eval: false

# Read CSV
read.csv("data.csv")
# Read Excel
readxl::read_excel("data.xlsx")
```
:::

::: {.column width="50%"}
**Writing Data:**

```{r}
#| echo: true
#| eval: false

# Save as CSV
write.csv(data, "output.csv")
# Save R objects
save(data, file = "data.RData")

```
:::
:::

## Data Visualization with ggplot2

```{r}
#| echo: true
#| eval: true
#| fig-width: 10
#| fig-height: 4

library(ggplot2)

# Sample data
sample_data <- data.frame(
  x = 1:10,
  y = c(2, 4, 3, 8, 6, 9, 7, 10, 12, 11),
  group = rep(c("A", "B"), each = 5)
)

# Create a professional plot
ggplot(sample_data, aes(x = x, y = y, color = group)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Sample Time Series Data",
       x = "Time", y = "Value", color = "Group") +
  theme(plot.title = element_text(hjust = 0.5, size = 14))
```

## Statistical Analysis Workflow {.fade-in-up}

::: {.tip-box}
**Reproducible Research Principles**: Document your analysis, use version control, and create reproducible scripts
:::

**Typical R Analysis Pipeline:**

1. **Data Import** → `read.csv()`, `read_excel()`
2. **Data Cleaning** → `dplyr` functions, missing value handling  
3. **Exploratory Analysis** → `summary()`, `cor()`, visualizations
4. **Statistical Modeling** → `lm()`, `glm()`, specialized packages
5. **Results Interpretation** → Coefficients, p-values, effect sizes
6. **Reporting** → R Markdown, Quarto, publication-ready outputs

## Next Steps and Resources

::: {.columns}
::: {.column width="50%"}
**Learning Path:**

- **Fundamentals**: R syntax, data types, functions
- **Data Manipulation**: tidyverse
- **Statistics**: Hypothesis testing, regression
- **Advanced Topics**: Shiny apps

:::

::: {.column width="50%"}
**Getting Help:**

```{r}
#| echo: true
#| eval: false

# Built-in help
?function_name
help("lm")

```

::: {.highlight-box}
**Remember**: R has a steep learning curve initially, but becomes incredibly powerful once you master the basics.
:::
:::
:::

---

**Questions?** 

Contact: your.email@institution.edu
